---
title: 'Activity 25: Predicates and the DE-9IM'
output-dir: docs
format:
   html:
    code-fold: true
    toc: true
date: "2025-05-05"
author: 
  name: Josh Puyear
  email: "puyearjosh@gmail.com" 
---
# 1. Download the data
# 2. Read the river shp file with read_sf
```{r, echo = TRUE}
library(AOI)
library(sf)
library(dplyr)
library(ggplot2)

rivshp <- read_sf("data/MajorRivers.shp")

```

# 3. Filter it to include only the Mississippi River system
v
# 4. Use AOI::aoi_get() to get the county boundaries for all counties in CONUS.
```{r, echo = TRUE}
mrb <- rivshp %>% 
  filter(SYSTEM == "Mississippi") %>% 
  st_as_sf(crs = 4326)#3

counties = aoi_get(state = "conus", county = "all")

county <- counties %>% 
  st_as_sf(crs = 4326)
```

# 5. Use st_filter to identify the counties that intersect the Mississippi River system
```{r, echo = TRUE}
#counties intersecting the Mississippi River system
ms_intersect <- st_filter(county, mrb) %>% 
  st_as_sf(crs = 4326)


```

# 6. Make a map of the counties that intersect the Mississippi River system, along with the rivers themselves.
```{r, echo = TRUE}
ggplot(counties) + 
  geom_sf() + 
  geom_sf(data = ms_intersect, fill = "blue", alpha = .3) +
  geom_sf(data = mrb, col = "darkblue", size = 50) +
  theme_void()

```
# 7 Read in the city data from last assignment/lab and ensure they are a sf object in the correct CRS
```{r, echo = TRUE}
uscity <- read_csv("data/uscities.csv")

#spatial feature of US cities
cities_sf <- st_as_sf(uscity, coords = c("lng", "lat"), crs = 4326) %>% mutate(point = "geometry")

```
# 8. Use st_join to identify the cities that intersect the counties that intersect the Mississippi River system
```{r, echo = TRUE}
ms_city_cty <- st_join(ms_intersect, cities_sf)
```

# 9. Calculate the total urban population in each county by summing the population of the cities within each county that intesects the Mississippi River system.
```{r, echo = TRUE}
ms_cty_pop <- ms_city_cty %>% 
  group_by(fip_code) %>%
  summarize(population = sum(population))
```

# 10. Add this data to the subset of intersecting counties and use it to color you map using scale_*
```{r, echo = TRUE}

ggplot() +
  # Base county layer (light background)
  geom_sf(data = county, fill = "gray90", color = NA) + 
  
  # Population-scaled counties
  geom_sf(data = ms_cty_pop, aes(fill = population), 
          color = "white", size = 0.1, alpha = 0.8) +
  
  # Mississippi River
  geom_sf(data = mrb, color = "darkblue", size = 0.5) +
  
  # Color scaling
  scale_fill_viridis_c(
    name = "population",
    option = "plasma",
    trans = "log10",  # Helps visualize wide population ranges
    labels = scales::comma  # Format numbers with commas
  ) +
  
  # Theme and labels
  theme_void() +
  labs(title = "Population Along the Mississippi River",
       subtitle = "Counties intersecting the river basin",
       caption = "Data: US Census Bureau")

```

11. Save your map as a png file in the images folder and submit it to Canvas.
```{r, echo = TRUE}
ggsave(last_plot(), filename = "msriver_county_pop.png", path = "C:/Users/Joshua Puyear/Documents/csu-undergrad/ess330-daily-exercises/imgs")
```