---
title: 'Activity 26: Introduction to Raster Data'
output-dir: docs
format:
   html:
    code-fold: true
    toc: true
date: "2025-05-07"
author: 
  name: Josh Puyear
  email: "puyearjosh@gmail.com" 
---
# Starting with this base, construct a VSI prefixed URL that allows rast to read the data directly from our class Github Repo.

```{r, echo = TRUE}
url6vsis3<- 'https:/vsis3/raw.githubusercontent.com/mikejohnson51/csu-ess-330/refs/heads/main/resources/foco-elev-cm.tif'

url6 <- 'https://raw.githubusercontent.com/mikejohnson51/csu-ess-330/refs/heads/main/resources/foco-elev-cm.tif'
```

# 2. Once the date is read in, describe the structure of the data in a few sentences:

```{r, echo = TRUE}
library(terra)

rast(url6)

```
The raster url leads to a Spatial Raster (single layer raster) with dimensions of 725 by 572 pixels. The extent is -769695, -752535, 1978485, 2000235, which are the coordinates in centimeters. the next row of rast() describes transforamtions done on the raster to fit a coordinate reference system specified projection, which is under the NAD83 datum to give a digital elevation map in centimeters of Fort Collins.


```{r, echo = TRUE}
url6vsi <- rast(glue::glue("/vsicurl/{url6}"))
```

# 3. Natively, the data has a horizonal unit of cm. Multiply the data by 0.0328084 to convert the data to feet.
```{r, echo = TRUE}
url6vsi_ft <- url6vsi * 0.0328084
```

# 4. Extract the values from the raster into a data.frame by using the dataframe=TRUE parameter in the values() function.
```{r, echo = TRUE}
foco_rast <- values(url6vsi_ft, dataframe = TRUE)
```

# 5. Use ggpubr to create a density plot of the elevation data in feet. Equit the plot with clear labs, and a title as well as a theme.
```{r, echo = TRUE}
library(ggpubr)

ggdensity(
  data = foco_rast,          # Use dataframe, not raster object
  x = "dem",               # Column containing elevation values
  fill = "lightgray",      # Fill color
  color = "black"          # Outline color
) +
  labs(
    title = "Density Plot of Fort Collins Elevation",
    x = "Elevation (feet)", # Correct axis label
    y = "Density"
  ) +
  theme_bw()

ggsave(filename = "focoelev_dens.png", plot = last_plot(), path = "imgs")

```

```{r, echo = TRUE}


```

```{r, echo = TRUE}


```

```{r, echo = TRUE}


```

```{r, echo = TRUE}


```

```{r, echo = TRUE}


```

```{r, echo = TRUE}


```

```{r, echo = TRUE}


```

```{r, echo = TRUE}


```

```{r, echo = TRUE}


```

```{r, echo = TRUE}


```

```{r, echo = TRUE}


```

```{r, echo = TRUE}


```

```{r, echo = TRUE}


```

```{r, echo = TRUE}


```

```{r, echo = TRUE}


```

```{r, echo = TRUE}


```

```{r, echo = TRUE}


```

```{r, echo = TRUE}


```