---
title: 'Activity 27: Mapping the Poudre'
output-dir: docs
format:
   html:
    code-fold: true
    toc: true
date: "2025-05-07"
author: 
  name: Josh Puyear
  email: "puyearjosh@gmail.com" 
---

Libraries
```{r, echo = TRUE}
library(sf)
library(osmdata)
library(crop)
library(raster)
library(terra)

```
Lecture Code


Define the AOI
```{r, echo = TRUE}
(bb = AOI::geocode("Fort Collins", bbox = TRUE) |> 
  st_transform(5070))
```

Read data from elevation map tiles, for a specific zoom, and crop to the AOI
```{r, echo = TRUE}

#data are inaccessible. wow.

#elev = rast('/vsis3/lynker-spatial/gridded-resources/dem.vrt') |> 
 # crop(bb)

#writeRaster(elev, filename = "data/foco-elev.tif", overwrite = TRUE)

#writeRaster(elev, filename = "../resources/foco-elev-cm.tif", overwrite = TRUE)
```

The resulting raster …
```{r, echo = TRUE}
(elev = rast("data/foco-elev.tif"))

```

Raster Values
```{r, echo = TRUE}

```

Raster Values
```{r, echo = TRUE}

```

All image files are the same!
```{r, echo = TRUE}

```

Raster Algebra
In these functions you can mix SpatRast objects with numbers, as long as the first argument is a raster object.

That means you can add 100 to a raster object but not a raster object to 100
```{r, echo = TRUE}

```


Replacement
Raster values can be replaced on a conditional statements
Doing this changes the underlying data!
If you want to retain the original data, you must make a copy of the base layer

```{r, echo = TRUE}

```

Modifying a raster
When we want to modify the extent of a raster we can clip it to a new bounds

crop: lets you reduce the extent of a raster to the extent of another, overlapping object:
```{r, echo = TRUE}

```


Modifying the underlying data:
mask: mask takes an input object (sf or SpatRast) and set anything not undelying the input to a new value (default = NA)
```{r, echo = TRUE}

```


What is mask doing?
```{r, echo = TRUE}

```


Crop or/and mask
Crop is more efficient then mask
Often you will want to mask and crop a raster
The correct way to do this is crop then mask
```{r, echo = TRUE}

```


Aggregate and disaggregate
aggregate and disaggregate allow for changing the resolution of a Raster object.

This is similar to the zoom scaling on a web map except the scale factor is not set to 2

For aggregate, you need to specify a function determining what to do with the grouped cell values (default = mean).
```{r, echo = TRUE}

```


app
Just like a vector, we can apply functions over a raster with app

These types of formulas are very useful for thresholding analysis

These types of functions are very simular in concept to map_*


```{r, echo = TRUE}

```

Create a conditional (threshold) mask
```{r, echo = TRUE}

```


Results
```{r, echo = TRUE}

```

Multiply cell-wise
```{r, echo = TRUE}

```

Reclassify
```{r, echo = TRUE}

```

Real Example: Classify Rainfall Regions of Colorado
```{r, echo = TRUE}

```

Plot
```{r, echo = TRUE}

```

Object/Field Interaction
Often, we want to know the relationship between a raster and a vector object
In these cases, we can use the extract function to get values of a raster at the locations (point or bounds) of a vector object
For this example, we will use OpenStreetMap data to get the location of rivers in Fort Collins.
```{r, echo = TRUE}
foco_rivers <- osmdata::opq(st_bbox(st_transform(fc,4326))) |> 
  add_osm_feature("waterway") |> 
  osmdata_sf()
```

Lets find the longest river segment IN our extent
```{r, echo = TRUE}
river = foco_rivers$osm_lines |> 
  st_transform(crs(r)) |> 
  st_intersection(st_as_sfc(st_bbox(r))) %>% 
  mutate(length = st_length(.)) |> 
  slice_max(length, n = 1)
```

```{r, echo = TRUE}
plot(r)
plot(river, add = TRUE, col = "blue", lwd = 2)
```


Value Extraction

Often, we want to know the profile and sinousity of a river
To do this, we need to know the inlet and outlet as well as the straight line connector
```{r,echo = TRUE}
inlet  = head(st_cast(river, "POINT"), 1)
outlet = tail(st_cast(river, "POINT"), 1) 
pts    = bind_rows(inlet, outlet) 

line = st_cast(st_union(pts), "LINESTRING")

plot(r)
plot(river, add = TRUE, col = "blue", lwd = 2)
plot(line, add = TRUE, col = "black", lwd = 2)
plot(outlet$geom, add = TRUE, pch = 16, col = "red")
plot(inlet$geom,  add = TRUE, pch = 16, col = "green")

```



#Assigment

1. To do use the code from lecture to extract all waterways from OpenStreetMap for the Bounding Box of Fort Collins, CO.

```{r, echo = TRUE}
fc = AOI::geocode("Fort Collins", bbox = TRUE)

foco_rivers <- osmdata::opq(st_bbox(st_transform(fc,4326))) |> 
  add_osm_feature("waterway") |> 
  osmdata_sf()

```

2. Filter the osm_lines object to only include the Cache la Poudre River and merge the lines into a single line object with st_union(). Be sure to convert the object to a sf object with st_as_sf() when done.

```{r, echo = TRUE}
library(sf)
library(dplyr)
library(stringr)

# Step 1: Extract osm_lines from the list and convert to sf
river_lines <- st_as_sf(foco_rivers$osm_lines)

# Step 2: Filter for Cache la Poudre (with name variations)
poudre_river <- river_lines %>% 
  filter(
    str_detect(tolower(name), "cache la poudre") |  # Primary name
    str_detect(tolower(name), "poudre river") |     # Common alternative
    (waterway == "river" & str_detect(tolower(name), "poudre"))  # OSM tag fallback
  )

# Step 3: Merge all segments into one line
poudre_merged <- poudre_river %>% 
  st_union() %>%             # Combine geometries
  st_as_sf() %>%             # Convert back to sf
  mutate(
    name = "Cache la Poudre River",
    length_m = st_length(.)  # Optional: add length attribute
  )

# Verify
plot(st_geometry(poudre_merged), col = "blue", lwd = 3)
```

3. Use st_length() to compute the length of the river for future calculations.

```{r, echo = TRUE}
library(sf)
library(dplyr)

# Assuming you've already created 'poudre_merged' from previous steps
poudre_merged <- poudre_merged %>%
  mutate(
    length_m = st_length(.),  # Calculate length in meters
    length_km = as.numeric(length_m) / 1000  # Convert to kilometers
  )

# Print the length
cat("Cache la Poudre River Length:",
    round(as.numeric(poudre_merged$length_m), 1), "meters",
    "(", round(poudre_merged$length_km, 1), "km)\n")

# Save for future use (as R object or CSV)
saveRDS(poudre_merged, "poudre_river_geometry.rds")  # Preserves geometry + attributes

# For tabular analysis:
poudre_length_data <- data.frame(
  name = "Cache la Poudre River",
  length_m = as.numeric(poudre_merged$length_m),
  length_km = poudre_merged$length_km
)
write.csv(poudre_length_data, "poudre_river_length.csv", row.names = FALSE)

```
4. Use st_cast() to convert the river object to a POINT object and save it as poudre_pts for latter extraction tasks

```{r, echo = TRUE}
library(sf)

# Convert the merged LINESTRING to POINTs (one point per vertex)
poudre_pts <- poudre_merged %>% 
  st_cast("POINT") %>%  # Convert geometry type
  mutate(
    point_id = 1:n(),  # Add unique ID to each point
    cumul_length_m = as.numeric(st_length(.)),  # Optional: cumulative length
    .before = 1  # Insert new columns at start
  )

# Verify
head(poudre_pts, 3)  # Check first few points
plot(st_geometry(poudre_pts), pch = 20, cex = 0.5, col = "red")  # Visual check

# Save for later use
saveRDS(poudre_pts, "poudre_river_points.rds")

```

5. Use the rast() function to read in the DEM file from the lynker-spatial S3 bucket shared in last assignment. Be sure to use the vsis3 prefix!

```{r, echo = TRUE}
library(terra)

url <- 'https://raw.githubusercontent.com/mikejohnson51/csu-ess-330/refs/heads/main/resources/foco-elev-cm.tif'

rast(url)


```

```{r, echo = TRUE}


```

```{r, echo = TRUE}


```

```{r, echo = TRUE}


```

```{r, echo = TRUE}


```

```{r, echo = TRUE}


```

```{r, echo = TRUE}


```